\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-03-26 main.tex
compare array with cubble, different formats spatial and temporal data are stored

@@

\caption{An illustration of the nested and long form in cubble. The nested form defines each site in a row and nests the time varying variables into a single column \code{ts}. The long form cubble uses \code{id} and \code{t} to identify each row and store the time invariant variables as an attribute, \code{spatial}.}\label{fig:illu-cubble}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_mig %>% tamp() %>% stretch()
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: long [dbl], lat [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\code{tsibble} is a temporal data structure that uses \code{index} and
\code{key} to identify the time and different series and \code{cubble}
can be seen as a natural extension of \code{tsibble} for spatio-temporal
data with an additional coordinates component and using two forms to
arrange variables. This makes it easy to cast a \code{tsibble} into a
\code{cubble} as only the \code{coords} argument needs to be supplied:

\begin{CodeChunk}
\begin{CodeInput}
R> # demonstrate with a tsibble created from climate_flat
R> raw <- climate_flat %>% 
+   tsibble::as_tsibble(key = id, index = date) 
R> 
R> dt <-  raw %>% 
+   cubble::as_cubble(coords = c(long, lat))
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> dt %>% 
+   mutate(fabletools::features(ts, tmax, list(tmax_mean = mean, tmax_var = var)))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      tmax_mean tmax_var
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>      <dbl>    <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_t~      25.7    38.6 
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_t~      26.2    51.1 
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_t~      23.7    45.4 
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_t~      33.1     3.02
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_t~      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

As a spatial data object, \code{sf} creates a simple feature geometry
list-column (\code{sfc}) in the data frame to provide spatial operations
on various geometry types (\code{POINT}, \code{LINESTRING},
\code{POLYGON}, \code{MULTIPOLYGON}, etc). These spatial operations are
also valuable for spatio-temporal data analysis, but an \code{sf} object
\emph{usually} don't contain temporal variables. This means \code{sf}
can't be directly cast into a \code{cubble}, however, \code{cubble} does
support \code{sfc} columns in the nested form and spatial operations
applied to the \code{sfc} column in \code{sf} can still be applied to
the \code{sfc} column in a cubble. The following example shows how to
create an \code{sfc} column of \code{POINT} type from latitude and
longitude in cubble. Then \code{sf::st_within} is used to add the state
\code{MULTIPOLYGON} of each weather station before a coordinate
transformation is made.

\begin{CodeChunk}
\begin{CodeInput}
R> library(sf)
R> # create a cubble
R> cb <- climate_flat %>% 
+   cubble::as_cubble(key = id, index = date, coords = c(long, lat))
R> 
R> aus <- ozmaps::abs_ste
R> 
R> dt <- cb %>%
+   mutate(
+     # create `sfc` column based on long and lat
+     ll = st_sfc(
+       purrr::map2(long, lat, ~st_point(c(.x, .y))),
+       crs = st_crs(aus)),
+     
+     # append state multi-polygon based on the `sfc` created
+     state = aus$geometry[st_within(ll, aus, sparse = FALSE)],
+     
+     # adopt a different projection: lambert conformal conic (EPSG:3112)
+     state = st_transform(state, crs = "EPSG:3112")
+     )
R> 
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name         wmo_id ts                        ll
  <chr>       <dbl> <dbl> <dbl> <chr>         <dbl> <list>           <POINT [°]>
1 ASN00009021 -31.9  116.  15.4 perth airpo~  94610 <tibble~ (115.9764 -31.9275)
2 ASN00010311 -31.9  117. 179   york          94623 <tibble~  (116.765 -31.8997)
3 ASN00010614 -32.9  117. 338   narrogin      94627 <tibble~ (117.1797 -32.9342)
4 ASN00014015 -12.4  131.  30.4 darwin airp~  94120 <tibble~ (130.8925 -12.4239)
5 ASN00015131 -17.6  134. 220   elliott       94236 <tibble~ (133.5407 -17.5521)
# ... with 1 more variable: state <MULTIPOLYGON [m]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(s2)
R> # western australia map
R> wa <- ozmaps::abs_ste %>% filter(NAME == "Western Australia")
R> 
R> # mutate a `s2_lnglat` vector on `cb` created in the last chunk
R> cb %>%
+   mutate(ll = s2_lnglat(long, lat)) %>% 
+   filter(s2_within(ll, wa))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [3]: nested form
# bbox:     [115.97, -32.94, 117.18, -31.89]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name          wmo_id ts                 ll      
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>             <s2_lng>
1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tibble [366 x 4]> (115.97~
2 ASN00010311 -31.9  117. 179   york           94623 <tibble [366 x 4]> (116.76~
3 ASN00010614 -32.9  117. 338   narrogin       94627 <tibble [366 x 4]> (117.17~
\end{CodeOutput}
\end{CodeChunk}

@@

Memory limit with NetCDF data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or a
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\code{long_range} and \code{lat_range} if the NetCDF file has finer
resolution than needed.

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], matched_var [dbl]
  id          name                lat  long type    dist group ts        n_match
  <chr>       <chr>             <dbl> <dbl> <chr>  <dbl> <int> <list>      <int>
1 405234      SEVEN CREEKS @ D~ -36.9  146. river   6.15     5 <tibble ~      21
2 404207      HOLLAND CREEK @ ~ -36.6  146. river   8.54    10 <tibble ~      21
3 ASN00082042 strathbogie       -36.8  146. clima~  6.15     5 <tibble ~      21
4 ASN00082170 benalla airport   -36.6  146. clima~  8.54    10 <tibble ~      21
5 230200      MARIBYRNONG RIVE~ -37.7  145. river   6.17     6 <tibble ~      19
6 ASN00086038 essendon airport  -37.7  145. clima~  6.17     6 <tibble ~      19
7 406213      CAMPASPE RIVER @~ -37.0  145. river   1.84     1 <tibble ~      18
8 ASN00088051 redesdale         -37.0  145. clima~  1.84     1 <tibble ~      18
\end{CodeOutput}
\end{CodeChunk}

>>>>

\hypertarget{conceptual-framework}{%
\section{Conceptual framework}\label{conceptual-framework}}

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by \citet{lu_multidimensional_2018} has discussed the
suitability of this representation through array algebra and practical
spatio-temporal tasks and software implementation has been seen in many
R packages: \pkg{raster} \citep{raster}, \pkg{terra} \citep{terra}, and
\pkg{stars} \citep{stars}.

While it is efficient to represent spatio-temporal data in an array,
there are situation where array {[}is not the best solution{]}.
Sometimes, one may want to create a new

Also, array is not the best solution for a large subset of data,
specifically, the vector spatio-temporal data. With vector data, spatial
and temporal variables are commonly collected in separate 2D tables and
there is usually a location ID to link the spatial and temporal
information. Sometimes, the geometrical shape of the location may not
present and an additional linking is needed to incorporate the map data.
With the multiple pieces in 2D data frame format, there should be a
solution to organise all these data from multiple sources in a 2D
structure, without needing to convert it into higher dimensional
representation.

Reliable software and data structure has been proposed in the spatial
and temporal domain, while a direct extension to spatio-temporal data is
not ideal. The main issue here lays in the format that spatial and
temporal data is stored. For spatial data, \pkg{sf} \citep{sf} covers a
wide range of geometrical operations and is the format most map data
stored. When incorporating the time dimension, a time wide format,
introduced in \citet{spacetime}, is usually adopted. This treats time as
variable name rather than value and makes it inconvenient to wrangle
time, as per the tidy data principle \citep{tidydata}. In the temporal
domain, \pkg{tsibble} \citep{tsibble} follows the tidy data principle to
arrange the time in the long format, while when adding spatial
variables, a long format implies repeating variables at each time and
this can make a huge difference when the repetition is on large
geometrical shapes with long time period.

However, with all the benefit of conceptualising the data into a cube,
the cubical structure is not particular convenient for, using the
tidyverse word, mutating new variables in the data object. This is
because an array is a higher dimensional object than the 2D data frame
and adding a new column doesn't clearly define how a vector should be
added to an (multidimensional) array.

The cubble package is a collection of tools to work with vector
spatio-temporal data from multiple sources.

@@

\caption{An illustration of the nested and long form in cubble. The nested form defines each site in a row and nests the time varying variables into a single column \code{ts}. The long form cubble uses \code{id} and \code{t} to identify each row and stores the time invariant variables as an attribute, \code{spatial}.}\label{fig:illu-cubble}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_mig %>% tamp() %>% stretch()
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\code{tsibble} is a temporal data structure that uses \code{index} and
\code{key} to identify the time and different series. \code{cubble} can
be seen as a natural extension of \code{tsibble} for spatio-temporal
data with an additional coordinates component and using two forms to
arrange variables. This makes it easy to cast a \code{tsibble} into a
\code{cubble} as only the \code{coords} argument needs to be supplied:

\begin{CodeChunk}
\begin{CodeInput}
R> # example with a tsibble created from climate_flat
R> raw <- climate_flat %>%
+   tsibble::as_tsibble(key = id, index = date)
R> 
R> dt <-  raw %>%
+   cubble::as_cubble(coords = c(long, lat))
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> dt %>%
+   mutate(fabletools::features(ts, tmax, list(tmax_mean = mean, tmax_var = var)))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name          wmo_id ts       tmax_mean tmax_var
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>       <dbl>    <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tbl_ts>      25.7    38.6 
2 ASN00010311 -31.9  117. 179   york           94623 <tbl_ts>      26.2    51.1 
3 ASN00010614 -32.9  117. 338   narrogin       94627 <tbl_ts>      23.7    45.4 
4 ASN00014015 -12.4  131.  30.4 darwin airpo~  94120 <tbl_ts>      33.1     3.02
5 ASN00015131 -17.6  134. 220   elliott        94236 <tbl_ts>      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

As a spatial data object, \code{sf} creates a simple feature geometry
list-column (\code{sfc}) in the data frame to provide spatial operations
on various geometry types (\code{POINT}, \code{LINESTRING},
\code{POLYGON}, \code{MULTIPOLYGON}, etc). These spatial operations are
also valuable for spatio-temporal data analysis, but an \code{sf} object
\emph{usually} does not contain temporal variables. This means \code{sf}
cannot be directly cast into a \code{cubble}, however, \code{cubble}
does support \code{sfc} columns in the nested form and spatial
operations applied to the \code{sfc} column in \code{sf} can still be
applied to the \code{sfc} column in a cubble. The following example
shows how to create an \code{sfc} column of \code{POINT} type from
latitude and longitude in cubble. Then \code{sf::st_within} is used to
add the state \code{MULTIPOLYGON} of each weather station before a
coordinate transformation is made.

\begin{CodeChunk}
\begin{CodeInput}
R> library(sf)
R> # create a cubble
R> cb <- climate_flat %>%
+   cubble::as_cubble(key = id, index = date, coords = c(long, lat))
R> 
R> aus <- ozmaps::abs_ste
R> 
R> dt <- cb %>%
+   mutate(
+     # create `sfc` column based on long and lat
+     ll = st_sfc(
+       purrr::map2(long, lat, ~st_point(c(.x, .y))),
+       crs = st_crs(aus)),
+ 
+     # append state multi-polygon based on the `sfc` created
+     state = aus$geometry[st_within(ll, aus, sparse = FALSE)],
+ 
+     # adopt a different projection: lambert conformal conic (EPSG:3112)
+     state = st_transform(state, crs = "EPSG:3112")
+     )
R> 
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name   wmo_id ts                        ll
  <chr>       <dbl> <dbl> <dbl> <chr>   <dbl> <list>           <POINT [°]>
1 ASN00009021 -31.9  116.  15.4 perth~  94610 <tibble> (115.9764 -31.9275)
2 ASN00010311 -31.9  117. 179   york    94623 <tibble>  (116.765 -31.8997)
3 ASN00010614 -32.9  117. 338   narro~  94627 <tibble> (117.1797 -32.9342)
4 ASN00014015 -12.4  131.  30.4 darwi~  94120 <tibble> (130.8925 -12.4239)
5 ASN00015131 -17.6  134. 220   ellio~  94236 <tibble> (133.5407 -17.5521)
# ... with 1 more variable: state <MULTIPOLYGON [m]>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(s2)
R> # Western Australia map
R> wa <- ozmaps::abs_ste %>% filter(NAME == "Western Australia")
R> 
R> # mutate a `s2_lnglat` vector on `cb` created in the last chunk
R> cb %>%
+   mutate(ll = s2_lnglat(long, lat)) %>%
+   filter(s2_within(ll, wa))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [3]: nested form
# bbox:     [115.97, -32.94, 117.18, -31.89]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name          wmo_id ts                 ll      
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>             <s2_lng>
1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tibble [366 x 4]> (115.97~
2 ASN00010311 -31.9  117. 179   york           94623 <tibble [366 x 4]> (116.76~
3 ASN00010614 -32.9  117. 338   narrogin       94627 <tibble [366 x 4]> (117.17~
\end{CodeOutput}
\end{CodeChunk}

@@

The memory limit with NetCDF data in cubble depends on the number of
longitude grid points times the number of latitude grid points times the
number of time grid points times the number of variables. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year. The spatial grid can be reduced to trade
for longer time periods and more variables. A 300 by 300 spatial grid
can be a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree
resolution or a global bounding box {[}-180, -90, 180, -90{]} at 1
degree resolution. Subsetting longitude and latitude grids is available
through \code{long_range} and \code{lat_range} if the NetCDF file has
finer resolution than needed.

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], matched_var [dbl]
  id          name                  lat  long type   dist group ts       n_match
  <chr>       <chr>               <dbl> <dbl> <chr> <dbl> <int> <list>     <int>
1 405234      SEVEN CREEKS @ D/S~ -36.9  146. river  6.15     5 <tibble>      21
2 404207      HOLLAND CREEK @ KE~ -36.6  146. river  8.54    10 <tibble>      21
3 ASN00082042 strathbogie         -36.8  146. clim~  6.15     5 <tibble>      21
4 ASN00082170 benalla airport     -36.6  146. clim~  8.54    10 <tibble>      21
5 230200      MARIBYRNONG RIVER ~ -37.7  145. river  6.17     6 <tibble>      19
6 ASN00086038 essendon airport    -37.7  145. clim~  6.17     6 <tibble>      19
7 406213      CAMPASPE RIVER @ R~ -37.0  145. river  1.84     1 <tibble>      18
8 ASN00088051 redesdale           -37.0  145. clim~  1.84     1 <tibble>      18
\end{CodeOutput}
\end{CodeChunk}
